The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-24 paper/ms.tex
small edits and rephrasings to section 3, and a few reference additions/changes

\section{Coordinate System Offsets I. Measurement Errors} \label{sec:mes_err}
We consider the first potential source of coordinate system offsets which
would generate the action errors described in Section~\ref{sec:ref_frame}:
errors in the determination of the Galactocentric parameters. Converting from
heliocentric to Galactocentric coordinates relies on these measurements.
Therefore, errors in their values will induce a systematic offset in the
Galactocentric phase-space position of any observed star.

\subsection{Galactic Center}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sgr~A\textsuperscript{*}
\citep[e.g.][]{2004ApJ...616..872R}. From stellar motions near
Sgr~A\textsuperscript{*}, the distance from the Sun to the Galactic center can
be measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}, with a recent
measurement using near-infrared interferometry of $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}.

We take the ``dynamical Galactic center'' to be the point in three-dimensional
space about which the stars in the solar neighborhood are orbiting. An
important point to make about the Galactic center is that while many authors
assume Sgr~A\textsuperscript{*} is colocated with the dynamical Galactic
center, this assumption has not yet been justified. One potential test is to
perform some sort of dynamical modelling where the distance from the Sun to
the Galactic center is a free parameter and can be constrained with real data.
For example, \citet{2015ApJ...803...80K} were able to make this measurement by
modelling the dynamics of the stream Palomar 5. Other dynamical measurements
of the Solar distance have been made, though none with a precision capable of
competing with the distance to Sgr~A\textsuperscript{*}
\citep{1981gask.book.....M, 2011PASJ...63..867S, 2012MNRAS.427..274S,
2013AstL...39...95B, 2013IAUS..289..444Z}.

\subsection{Galactic Orientation}
Second, one must define the orientation of the Galaxy about the center. This
was defined in 1958 by the IAU subcomission 33b \citep{1960MNRAS.121..123B} by
defining the coordinates of the Galactic center in B1950 coordinates as
(17:42:26.6, -28:55:00) and the north Galactic pole as (12:49:00, +27:24:00).
However, there is growing evidence that the true midplane is tilted when
considering stars \citep{2014ApJ...797...53G, 2016ARAA..54..529B}, though not
with H~\textsc{ii} \citep{2019ApJ...871..145A}.

\subsection{Solar Height}
Thirdly, one must define the Sun's vertical distance from the Galactic
midplane, which can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars). The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from {\em Gaia} DR2
placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another strategy is
to use the cold gas or H~\textsc{ii} regions in the disk to define the Galactic
midplane, leading to slightly different values ($\sim 5\,\pc$) for the Sun's
relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review
of these measurements is given by \citet{2016ARAA..54..529B}.

\subsection{Local Standard of Rest}
Finally, one must define the local standard of rest (LSR). The radial
($U_{\odot}$) and vertical ($W_{\odot}$)components are easily computed by
taking the mean motions of different stellar groups. The azimuthal component
($V_{\odot}$) is more difficult to measure, but can be modelled using the
asymmetric drift relation \citep{2008gady.book.....B}. The values of the
components of the LSR are usually taken from \citet{2010MNRAS.403.1829S}. The
value of the circular velocity is taken to be $\sim 220\kms$
\citep[e.g.][]{2012ApJ...759..131B}.

In order to convert a star's position and velocity from heliocentric to
Galactocentric coordinates, a number of parameters must be measured. Given the
great interest in each parameter, considerable effort has been placed on these
measurements. However, uncertainties remain, and detailed dynamical modelling
across the disk --- particularly for dynamically cold stars --- may have to
take them into account. For instance, the discrepancy of $\sim 5\,\pc$ between
the midplane defined by stars and by gas will induce a $\sim10\%$ error in
$J_z$ for an orbit with $z_{\text{max}}\sim100\,\pc$ (see
Section~\ref{sec:ref_frame}).

@@

We now consider a second source of systematic offsets in the Galactocentric
coordinate system --- intrinsic variations of the midplane across the disk. We
study this in two sets of simulations of Milky Way-mass galaxies. We will also
briefly discuss a rough estimate of the midplane variation derived from the
observed velocity distribution of the Milky Way, but defer a detailed
treatment to future work.

@@

\subsection{Description of Sagittarius Simulaton} \label{ssec:sag_sim}
In addition to the cosmological zoom-in, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modelled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modelled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modelled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$.

>>>>

\section{Coordinate System Offsets I. Measurement Errors} \label{sec:mes_err}
We consider the first potential source of coordinate system offsets which
would generate the action errors described in Section~\ref{sec:ref_frame}:
uncertainties in measurements of the position and velocity of the Sun relative to the Galactic center. Converting from
heliocentric to Galactocentric coordinates relies on these measurements.
Therefore, errors in their values will induce a systematic offset in the
Galactocentric phase-space position of any observed star.

\subsection{Galactic Center}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sgr~A\textsuperscript{*}
\citep[e.g.][]{2004ApJ...616..872R}. From stellar motions near
Sgr~A\textsuperscript{*}, the distance from the Sun to Sgr~A\textsuperscript{*} can
be precisely measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}, with a recent
measurement using near-infrared interferometry of $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}.

Importantly, the location of Sgr~A\textsuperscript{*} may not be equivalent to the location of the ``dynamical Galactic center:'' the point in three-dimensional
space about which the stars in the solar neighborhood are orbiting. This assumption, although frequently made, has not yet been justified. It is tested in any construction of a  dynamical model where the distance from the Sun to
the Galactic center, $R_0$, is a free parameter.
For example, \citet{2015ApJ...803...80K} measured $R_0$ while
modelling the dynamics of the stream Palomar 5. Many other dynamical measurements
of $R_0$ have been made (\citealt{2016ARAA..54..529B} summarize many pre-{\em Gaia} results) but none have yet achieved a precision comparable to that of the distance to Sgr~A\textsuperscript{*}.
% \citep{1981gask.book.....M, 2011PASJ...63..867S, 2012MNRAS.427..274S,2013AstL...39...95B, 2013IAUS..289..444Z}.

\subsection{Galactic Orientation}
Second, one must define the angular orientation of the Galaxy. This
was defined in 1958 by the IAU subcomission 33b \citep{1960MNRAS.121..123B} by
defining the coordinates of the Galactic center in B1950 coordinates as
(17:42:26.6, -28:55:00) and the north Galactic pole as (12:49:00, +27:24:00). These two quantities, together with $R_0$, define the orientation of the Galactic plane.
However, there is growing evidence that the stellar midplane is tilted relative to this coordinate system \citep{2014ApJ...797...53G, 2016ARAA..54..529B}, though not
the H~\textsc{ii} midplane \citep{2019ApJ...871..145A}.

\subsection{Solar Height}
Third, one must define the Sun's vertical distance from the Galactic
midplane, which can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars). The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from {\em Gaia} DR2
placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another strategy is
to use the cold gas or H~\textsc{ii} regions in the disk to define the Galactic
midplane, leading to slightly different values ($\sim 5\,\pc$) for the Sun's
relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review
of these measurements is given by \citet{2016ARAA..54..529B}.

\subsection{Local Standard of Rest}
Finally, one must define the local standard of rest (LSR), or mean velocity of stars near the Sun relative to the Galactic center (which is defined to have zero velocity), and the velocity of the Sun relative to the LSR. The radial
($U_{\odot}$) and vertical ($W_{\odot}$) components are computed by
taking the mean motions of different stellar groups \citep[e.g.][]{2012MNRAS.427..274S}. The azimuthal component
($V_{\odot}$) is more difficult to measure, but can be modelled using the
asymmetric drift relation \citep{2008gady.book.....B}. The values of the
components of the LSR are usually taken from \citet{2010MNRAS.403.1829S}. The
value of the circular velocity is taken to be $\sim 220\kms$
\citep[e.g.][]{2012ApJ...759..131B}.

\vspace{6pt}

Considerable effort has been placed on each of these
measurements, but uncertainties remain, and detailed dynamical modelling
across the disk --- particularly for dynamically cold stars --- may have to
take them into account. For instance, the discrepancy of $\sim 5\,\pc$ between
the midplane defined by stars and by gas will induce a $\sim10\%$ error in
$J_z$ for an orbit with $z_{\text{max}}\sim100\,\pc$ (see
Section~\ref{sec:ref_frame}).

@@

We now consider a second source of systematic offsets in the Galactocentric
coordinate system: intrinsic variations of the midplane across the disk. We
use two sets of simulations of Milky Way-mass galaxies to estimate the size of this effect. We will also
briefly discuss a rough estimate of the midplane variation derived from the
observed velocity distribution of the Milky Way, but defer a detailed
treatment to future work.

@@

\subsection{Description of MW-Sagittarius Interaction Simulation} \label{ssec:sag_sim}
In addition to the cosmological zoom-in, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modeled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modeled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modeled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All components are realized with live distributions of N-body particles; the MW and Sgr are each initialized to be in equilibrium in isolation \Gus{check!}.
